---
title: "DFO Reference Points 101 (Revised draft)"
output: html_document
author: "Tim Barrett and Robyn Forrest"
date: "November 2022"
---

```{r setup, include=FALSE}
library(ggplot2)
library(SAMtool)
knitr::opts_chunk$set(echo = T, warnings= F)
```

## Exercise 4: Reference points using an age-structured model with time-varying growth

***DISCLAIMER: These are illustrative, deterministic models and should not be used for stock assessment***

**Run the R chunks in Rstudio using the green arrow in each chunk. Run individual lines in the chunks using ctrl+enter. If you want to explore further, you can add your own R code inside any of the chunks, or add your own chunks using ctrl+alt+i. Click on the Knit button in Rstudio to create the document.**

Here we look at one operating model from Exercise 3 where growth is time-varying. The objective of this exercise is to explore the influence of the time period selected to describe growth on reference points and to explore *F* and *SSB*~MSY~ proxy reference points based on Spawning Potential Ratio. 

An operating model was conditioned to age-composition data for the catch and a survey index for a herring stock (time series of 53 years). An R object is provided for the operating model that contains the relevant data for the exercise. The model was conditioned using a Beverton-Holt stock-recruit relationship (SHH) with assumed steepness (*h*) = 0.8 over the first 5 years, and a constant *M* = 0.3 for all years and ages. The SRR is plotted below with model estimated Beverton-Holt parameters. For this exercise we assume that selectivity is equal to the maturity ogive and is constant over all years. Weight-at-age varies annually and here we look at how the changes in weight-at-age influence *SSB*~0~ and *SSB*~MSY~. We also look at comparing *F*~40%SPR~ to *F*~MSY~ and the equilibrium SSB from fishing at *F*~40%SPR~ as a proxy for *SSB*~MSY~.

```{r,warning=F,message=F}
source("source_ftns.r")
datalist <- readRDS("Ex4data.rda")
# The DF dataframe has the data needed for this exercise
DF <- datalist$DF
head(DF)
# Weight-at-age is time-varying and a matrix of dimension 12 x 53 is provided with the weights-at-age 0 to 11+ for 53 years
WAA <- datalist$WAA
WAA[,1:5]

```

Plot of weight-at-age over time:

```{r,warning=F}
library(reshape2)
WAADF <- melt(WAA)
colnames(WAADF)<-c('Age','Year','Weight')

ggplot(WAADF)+geom_path(aes(y=Weight,x=Year,colour=as.factor(Age))) + theme_classic() 
```

## Influence of changes in weight-at-age on $SSB_0$, and $SSB_{MSY}$

We first explore the influence of using the average weight-at-age from 

a) the first 5 years of the time series, and 
b) the last 5 years of the time series

on estimates of the unfished SSB-per-recruit ($\phi$~0~), $SSB_0$, and $SSB_{MSY}$.

```{r,warning=F,message=F}

surv0 <- survivorship_F(M=DF$M[1:12],n_ages=12)

# SSB-per-recruit (g/rec)
phi0_first5 <- sum(surv0*apply(WAA[,1:5],1,mean)*DF$mat[1:12])
phi0_first5
phi0_last5 <- sum(surv0*apply(WAA[,49:53],1,mean)*DF$mat[1:12])
phi0_last5
# SSB0
SSB0_first5 <- (DF$BHa[1]*phi0_first5-1)/DF$BHb[1] 
SSB0_last5 <- (DF$BHa[1]*phi0_last5-1)/DF$BHb[1] 

ggplot(DF)+geom_point(mapping=aes(y=Rec,x=SSB)) + theme_classic() +
  labs(x="SSB (kt)",y="Age 0 Recruitment in millions") +
  scale_x_continuous(limits = c(0,1200), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,12), expand = c(0, 0)) +
  geom_function(fun=function(x) (DF$BHa[1]*x/(1+DF$BHb[1]*x))) +
  geom_function(fun=function(x) (1/phi0_first5*x),colour="red",linetype=2) +
  geom_function(fun=function(x) (1/phi0_last5*x),colour="purple",linetype=2) +
  geom_vline(xintercept=SSB0_first5,color="red") +
  geom_vline(xintercept=SSB0_last5,color="purple") 

```

##Delete - I did want to show than h changes over time to demonstrate that h is not constant but is a measure of producivity and varies over time as components of productivity change. Too complicated? Or include?

```{r,warning=F,message=F}

RP_first5 <- MSYcalc(M=DF$M[1:12],waa=apply(WAA[,1:5],1,mean),mat=DF$mat[1:12],sel=DF$sel[1:12],a=DF$BHa[1],b=DF$BHb[1])
RP_last5 <- MSYcalc(M=DF$M[1:12],waa=apply(WAA[,49:53],1,mean),mat=DF$mat[1:12],sel=DF$sel[1:12],a=DF$BHa[1],b=DF$BHb[1])

ggplot(DF)+geom_point(mapping=aes(y=Rec,x=SSB)) + theme_classic() +
  labs(x="SSB (kt)",y="Age 0 Recruitment in millions") +
  scale_x_continuous(limits = c(0,1200), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,12), expand = c(0, 0)) +
  geom_function(fun=function(x) (DF$BHa[1]*x/(1+DF$BHb[1]*x))) +
  # SSB0 solid vertical line
  geom_vline(xintercept=SSB0_first5,color="red") +
  geom_vline(xintercept=SSB0_last5,color="purple") +
  # SSBmsy dashed vertical line
  geom_vline(xintercept=RP_first5$SSBmsy,color="red",linetype=2) +
  geom_vline(xintercept=RP_last5$SSBmsy,color="purple",linetype=2) 

```

## MSY proxies: 40% SPR

*F*~40%SPR~ is a common proxy for *F*~MSY~ for stocks with average productivity. Here we estimate *F*~40%SPR~ and the equilibrium SSB from fishing at *F*~40%SPR~ using the weight-at-age data from the last 5 years. The output of the `MSYcalc()` function provides a dataframe we call **RP** below that contains a vector of *F* and the SSB-per-recruit, YPR, equilibrium recruitment, and equilibrium SSB at each *F*.

```{r,warning=F,message=F}

RP <- RP_last5$DF

# Preview dataframe 
head(RP)

# Add SPR column to RP
RP$SPR <- RP$phi_f/RP$phi_f[1]

# Find the F where SPR is closest to 40%
Fspr40 <- RP$f[which(abs(RP$SPR-0.4)==min(abs(RP$SPR-0.4)))]

# Compare Fspr40 to Fmsy
Fspr40
RP_last5$Fmsy

# The SPR at Fmsy is estimated to be closer to 35%
RP$SPR[RP$f==round(RP_last5$Fmsy,3)]

# The equilibrium SSB from fishing at Fspr40 is a proxy for SSBmsy
RP$eq_ssb_f[RP$f==Fspr40]
RP_last5$SSBmsy
```


