---
title: "TESA Reference Points 101"
output: html_document
author: "Tim Barrett and Robyn Forrest"
date: "November 15 and 22, 2022"
---

```{r setup, include=FALSE}
library(tidyverse)
library(reshape2)
source("source_ftns.r")
knitr::opts_chunk$set(echo = T, warnings= F)
```

## Exercise 2: Estimating unfished biomass and MSY reference points using an age-structured model with a Beverton-Holt stock-recruit relationship

***DISCLAIMER: These are illustrative, deterministic models and should not be used for stock assessment***

**Run the R chunks in Rstudio using the green arrow in each chunk. Run individual lines in the chunks using ctrl+enter. If you want to explore further, you can add your own R code inside any of the chunks, or add your own chunks using ctrl+alt+i. Click on the Knit button in Rstudio to create the document.**

Here, we will extend the ideas from Exercise 1 to an age-structured model. By the end of the exercise you will learn how to:

* Calculate equilibrium unfished biomass ($B_\text{0}$) and equilibrium MSY reference points ($B_\text{MSY}$, $F_\text{MSY}$, $\text{MSY}$) when a stock-recruitment relationship is assumed

We will use outputs based on an example operating model that has been conditioned on age-composition data, catch and a survey index for a cod stock for ages 4+.

We assume a Beverton-Holt stock-recruit relationship (Beverton and Holt 1957; Hilborn and Walters 1992, Ch 7). The stock-recruit pairs are plotted below and saved in the dataframe **SR**. The biological data (weight-at-age, maturity-at-age, *M*-at-age) and assumed selectivity-at-age are saved in the dataframe **BI**.

```{r,warning=F, fig.show="hold", out.width="75%"}
data <- readRDS("Ex2data.rda") # data is a list with model outputs and data

# Dataframe SR with Year, SSB in kt, and age 0 recruitment in millions
SR <- data$SR

# Dataframe BI with ages (0 to 12+), weight-at-age in kg (waa), proportion mature-at-age (mat), M-at-age, selectivity-at-age (sel)
BI <- data$BI

# The Beverton-Holt stock-recruit relationship is parameterized in terms of R0, h, and phi0 [R0 and h are model estimated and phi0 is determined from biological parameters: M-at-age, weight-at-age, and maturity-at-age]. These parameters are stored in the data list:
h <- data$h
R0 <- data$R0
phi0 <<- data$phi0

# Preview data
head(SR)
head(BI)

# Plot the stock-recruitment relationship
P <- ggplot(SR) + geom_point(mapping=aes(y=Rec,x=SSB,colour=Year)) + theme_classic() + 
  labs(x="SSB (kt)",y="Age 0 Recruits (millions)") +
  scale_x_continuous(limits = c(0,8), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,40), expand = c(0, 0)) +
  geom_function(fun=function(x) (4*R0*h*x/((1-h)*R0*phi0+(5*h-1)*x)),colour="black",linetype=1) +
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold"))
P
```

<i><b>Figure 1. Model estimated recruitment (age-0 recruits in millions) vs. spawning stock biomass (in kt) with the stock-recruit relationship (SRR; black line)</i></b>

Age-structured models keep track of the number of fish that are alive at a given age and time (determined by recruitment, natural mortality and fishing mortality), the proportion mature at a given age (determined by the maturity-at-age schedule), as well as the proportion of fish that are vulnerable (or "available") to the fishery (determined by the vulnerability-at-age schedule). For this exercise, we assume these schedules are fixed in time. 

```{r,warning=F, fig.show="hold", out.width="75%"}

# Plot maturity-at-age and selectivity-at age
matsel <- BI %>% 
  select(age,mat,sel) %>%
  melt(id="age", variable.name="Schedule",value.name="Proportion") %>%
  # Rename labels mat, sel and age for prettier plotting
  mutate(Schedule=as.character(Schedule)) %>% 
  mutate(Schedule = replace(Schedule, Schedule=="mat","Maturity"),
         Schedule = replace(Schedule, Schedule=="sel","Vulnerability")) %>% 
  rename("Age"=age) %>%
  ggplot() + 
  geom_line(mapping=aes(x=Age,y=Proportion,colour=Schedule), size=1.5) + 
  theme_classic() + 
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold"))+
  theme(legend.title = element_text(size=12, face="bold"))+
  theme(legend.title = element_text(size=12))+
  scale_x_continuous(limits = c(0,12), breaks = 0:12, expand = c(0, 0)) 
matsel


```

<i><b>Figure 2. Maturity ogive and vulnerability ogive</i></b>


In **Figure 2**, we can see that the vulnerability curve is to the right of the maturity curve, implying that fish mature and have a chance to spawn before they are vulnerable to the fishery. 

Note that the terms vulnerability and selectivity are used somewhat interchangeably in fisheries. The term vulnerability captures both the "availability" of fish to the gear  (which could be determined by the location of the fishery relative to demographic components of the population, e.g., patterns in depth distribution) and "selectivity" of the fishing gear (e.g., mesh size, hook size or escape panels). However, the term selectivity is in common use as an interchangeable term with vulnerability. In these exercises, we will try to use the term vulnerability to be consistent with the equations but will sometimes use the term selectivity.

## Equilibrium Calculations

Now we look at the equilibrium calculations used to calculate reference points, given a set of model parameters. These exercises use deterministic parameter values but please be aware that stock assessments estimate parameters with uncertainty and that this uncertainty carries through into the reference points.

### Unfished Per-recruit Calculations

As for the simpler surplus production model in the previous exercise, the equilibrium unfished biomass ($B_0$) is the biomass that would result from running out the population model for a long period without fishing under constant average conditions. Usually reference points are based on spawning biomass (SSB), as this is the component of the population of most conservation interest and the equilibrium unfished biomass is $SSB_0$.

However, we can calculate equilibrium model behaviour more efficiently than running out models for many years, using equilibrium "per-recruit" methods, or "Botsford's incidence functions" (see Walters and Martell 2004, Box 3.1). The equations are similar to those used in the time-dynamic population model, but are based on the idea of survivorship-at-age ($l_a$), which tracks the proportion of recruits that survive from age to age. In the absence of fishing (i.e., $F$ = 0), the unfished equilibrium survivorship is given by:

\begin{equation*}
\tag{1}
  l_{0_{a}} = \left\{ 
  \begin{matrix}
    1 & , & a = a_{rec} \\ 
    l_{0_{a-1}}e^{-(M_{a-1})} & , & 1 < a_{rec} < a_{max} \\ 
    \frac{l_{0_{a-1}}e^{-(M_{a-1})}}{1-e^{-(M_{a})}} & , & a = a_{max}
  \end{matrix}
  \right\}
\end{equation*}

where $a$ = age, $a_{rec}$ = age-at-recruitment, $a_{max}$ = maximum age (plus group), $M_{a}$ = natural mortality rate-at-age, and $l_{0_{a}}$ is 1, by definition, at $a_{rec}$.

The equilibrium unfished SSB-per-recruit ($\phi_{E0}$) can then be calculated as the sum-product of the unfished equilibrium survivorship-at-age ($l_{0_{a}}$), the weight-at-age ($w_{a}$), and the maturity-at-age ($m_{a}$):

\begin{equation}
\tag{2}
{\phi_{E0}} = \sum_{a=a_{rec}}^{a_{max}} l_{0_{a}}w_{a}m_{a}
\end{equation}

where we use the subscript $_{E}$ because we assume that SSB is a proxy for eggs produced. Note that $\phi_{E0}$ here is the same as the $\phi_{0}$ parameter defined in the stock-recruit relationship when biological parameters (i.e., $M$, weight-at-age, and maturity-at-age) are not time-varying. We obtain an estimate of the equilibrium unfished SSB ($SSB_0$) from $\phi_{E0}$ and the unfished recruitment $R_{0}$ as: 

\begin{equation}
\tag{3}
{SSB_0} = {\phi_{E0}}{R_0}
\end{equation}

*Note:* The `survivorship_F()` function in source_ftns.R can be used here to calculate $l_a$ for any $F_e$ as shown in Equation 4. To obtain unfished $l_{0_{a}}$, set f = 0 in the `survivorship_F()` function.

```{r,message=F}
surv0 <- survivorship_F(f=0,M=BI$M,n_ages=nrow(BI)) # This is Equation 2 at F = 0
phiE0 <- sum(surv0*BI$waa*BI$mat) # This is Equation 1 at F = 0
phiE0 # unfished SSB-per-recruit

# R0 is an estimated parameter from the stock-recruit relationship
SSB0 <- phiE0*R0 # SSB0 = unfished SSB-per-recruit * unfished number of recruits
SSB0
```

The estimation of $SSB_0$ can be shown visually as the SSB at the intersection of the SRR and a line through the origin with slope equal to the inverse of $\phi_0$ (i.e., unfished recruits per spawner). This is illustrated below:

```{r,out.width="75%"}
P + geom_function(fun=function(x) (1/phiE0*x),colour="red") +
    geom_vline(xintercept=SSB0, linetype=2, color = "red") 
```

<i><b>Figure 2. Model estimated recruitment (age-0 recruits in millions) vs. spawning stock biomass (in kt) with the stock-recruit relationship (SRR; black line), and a line with slope $\frac{1}{\phi_{E0}}$ (red line) that intersects the SRR at $SSB_0$ (vertical dashed line)</i></b> 

### Per-recruit calculations with fishing

In order to calculate MSY-based reference points, we also need to calculate long-term equilibrium population behaviour under constant fishing mortality rates. At any constant equilibrium fishing mortality rate $F_{e}$, the equilibrium survivorship-at-age $l_a$ is a simple modification of **Equation 1**, accounting for fishing mortality and vulnerability-at-age $v_{a}$:

\begin{equation*}
\tag{4}
  l_{a} = \left\{ 
  \begin{matrix}
    1 & , & a = a_{rec} \\ 
    l_{a-1}e^{-(M_{a-1}+F_{e}v_{a-1})} & , & 1 < a_{rec} < a_{max} \\ 
    \frac{l_{a-1}e^{-(M_{a-1}+F_{e}v_{a-1})}}{1-e^{-(M_{a}+F_{e}v_{a})}} & , & a = a_{max}
  \end{matrix}
  \right\}
\end{equation*}

SSB-per-recruit ($\phi_{F}$) is then given by

\begin{equation}
\tag{5}
{\phi_{F}} = \sum_{a=a_{rec}}^{a_{max}} l_{a}w_{a}m_{a}
\end{equation}

### Deriving the Stock-recruit Parameters Needed for Calculating Equilibrium SSB and Equilibrium Recruitment

To calculate the MSY reference points, we can use the Beverton-Holt stock-recruit function (see footnote for Ricker equations) to calculate the equilibrium SSB and equilibrium recruitment for any $F_e$. We will focus on the Beverton-Holt stock-recruit function parameterized by two parameters $\alpha$ and $\beta$:

\begin{equation*}
\tag{6}
  R = \frac{\alpha SSB}{1+\beta SSB}
\end{equation*}

The $\alpha$ parameter represents the maximum juvenile survival rate, while the $\beta$ parameter determines the degree of compensation in juvenile survival. Note that we can re-express **Equation 6** in terms of SSB-per-recruit, by substituting $SSB=\phi_{F_e}R_e$ into **Equation 6** and rearranging to get and expression for $R_0$ in terms of $\alpha$, $\beta$, and $\phi_{E0}$:

\begin{equation*}
\tag{7}
R_0 = \frac{\alpha \phi_{E0}-1}{\beta \phi_{E0}} 
\end{equation*}

Our model has estimated the leading parameters $R_0$ and steepness $h$ using the parameterization:

\begin{equation*}
\tag{8}
  R = \frac{4 R_0 h SSB}{(1-h)R_0\phi_0 + (5h-1)SSB}
\end{equation*}

The $\alpha$ and $\beta$ parameters can be calculated from $R_0$, $h$, and $\phi_0$ as:

\begin{equation*}
\tag{9}
  \alpha = \frac{4h}{\phi_{0} (1-h)}
\end{equation*}

For more on derivation of $\alpha$, see Forrest et al. 2010 and references therein. Once we have an estimate of $\alpha$, we can obtain $\beta$ by rearranging Equation 7:

\begin{equation*}
\tag{10}
  \beta = \frac{\alpha \phi_{0}-1}{R_0 \phi_{E0}}
\end{equation*}

```{r}
# Calculate the BH parameters alpha and beta
alpha <- 4*h/(phi0*(1-h))
beta <- (alpha*phi0-1)/(R0*phi0)
```

## Calculate the MSY reference points ($F_\text{MSY}$, $SSB_\text{MSY}$, $MSY$)

To calculate $SSB_\text{MSY}$ (i.e., $B_\text{MSY}$), we need an estimate of equilibrium recruitment at $F_\text{MSY}$. To identify $F_\text{MSY}$, we therefore need to calculate equilibrium recruitment $R_e$ over a range of equilibrium fishing mortality rates $F_e$. We saw in **Equation 3** that we could estimate $SSB_{0}$ from an estimate of $\phi_{E0}$ and $R_{0}$. Therefore, we can estimate equilibrium SSB ($SSB_{e}$) from $\phi_{F}$ and equilibrium recruitment ($R_{e}$), where $\phi_{F}$ is evaluated at any alternative value of $F$.

We calculate $R_{e}$ for any given $F_{e}$ following Equation 7 with $\phi_{F}$ evaluated at the value $F_{e}$, i.e., 

\begin{equation*}
\tag{11}
R_{e} = \frac{\alpha \phi_{F}-1}{\beta \phi_{F}} 
\end{equation*}

and

\begin{equation*}
\tag{12}
SSB_e = \phi_{F} R_{e}
\end{equation*}

$F_\text{MSY}$ is the fishing mortality rate that maximizes the yield over the long term, where yield-per-recruit is given by a per-recruit version of the Baranov equation:

\begin{equation}
\tag{13}
{YPR_{e}} = \sum_{a=a_{rec}}^{a_{max}}       l_{a}w_{a}(1-e^{-(M_a+F_ev_a)})\frac{F_ev_a}{M_a+F_ev_a}
\end{equation}

then long-term equilibrium yield $Y_e$ is simply the product of the yield-per-recruit and the number of recruits:

\begin{equation*}
\tag{14}
Y_e = YPR_{e} R_e 
\end{equation*}

Unlike the surplus production model, there is no analytical solution for $F_\text{MSY}$. To find the value of $F_e$ that maximizes **Equation 14**, we need to do a numerical search. Usually in stock assessments this is done using an efficient optimization algorithm that searches for the value of $F_e$ that gives $\frac{\delta Y}{\delta F} = 0$. Here we use a simple numerical grid search over a vector of values of $F_e$, as this most clearly shows the mechanism of finding $F_\text{MSY}$, but talk to your colleagues about the algorithms they use in their stock assessments. 

*Note: If you are interested in this topic, there are some papers that show that if you flip the direction of estimation in the stock assessment model, i.e., if you estimate $F_\text{MSY}$ and $MSY$ as leading parameters instead of $R_0$ and $h$, then there actually is an analytical solution for the recruitment parameters (Schnute and Kronlund 1996; Forrest et al. 2008; Martell et al. 2008). However, we will not discuss this approach further in this exercise.*

The `MSY_calc()` function in source_ftns.R can be used to find $F_\text{MSY}$, $B_\text{MSY}$ and $MSY$, given the stock recruit parameters \alpha and \beta and the biological data. The calculations in this function are shown below and follow these steps:

1) Define a vector for $F_e$ `f <- seq(0,5,0.001)`;

2) For each $F_e$, calculate the SSB-per-recruit ($\phi_{F}$; Equation 5), the equilibrium recruitment at $F_e$ ($R_e$; Equation 11), and the yield-per-recruit at $F_e$ ($YPR_{e}$; Equation 13);

3) For each $F_e$, calculate the equilibrium yield $Y_e$ as the product of $YPR_F$ and $R_e$ at $F_e$ (Equation 14); 

4) Identify $F_{MSY}$ as the value of $F_e$ that maximizes $Y_e$;

5) Calculate $SSB_{MSY}$ as the product of $\phi_{F}$ and $R_e$ evaluated at $F_{MSY}$ (Equation 12);

6) Calculate $MSY$ as $Y_e$ evaluated at $F_{MSY}$.


```{r,message=F}
# 1)
f <- seq(0,5,0.001) # vector of fishing mortality rates

# Define a dataframe (DF) with columns for f, SSB-per-recruit (phi), yield-per-recruit, equilibrium SSB, equilibrium recruitment, and yield
DF <- data.frame(f=f, phi_f=NA, ypr_f=NA, eq_ssb_f=NA, eq_rec_f=NA, yield_f=NA)

# 2)
# SSB-per-recruit
for(i in 1:length(f)){
 DF$phi_f[i] <- sum(survivorship_F(f=f[i],M=BI$M,n_ages=length(BI$sel),sel=BI$sel,message=F)*BI$waa*BI$mat)
}  

# Equilibrium recruitment
DF$eq_rec_f <- (alpha*DF$phi_f-1)/(beta*DF$phi_f) # Equation 11

# Yield-per-recruit
for(i in 1:length(f)){
  DF$ypr_f[i] <- sum(survivorship_F(f=f[i],M=BI$M,n_ages=length(BI$sel),sel=BI$sel,message=F)*BI$waa*
                 (1-exp(-(BI$M+f[i]*BI$sel)))*f[i]*BI$sel/(BI$M+f[i]*BI$sel))
} 

# 3) Equilibrium yield

DF$yield_f <- DF$ypr_f*DF$eq_rec_f # Equation 14  

# 4)
Fmsy <- f[which(DF$yield_f == max(DF$yield_f))] # F that maximizes yield
Fmsy

# 5) Equilibrium SSB
DF$eq_ssb_f <- DF$eq_rec_f*DF$phi_f # Equation 12

SSBmsy <- DF$eq_ssb_f[which(DF$yield_f == max(DF$yield_f))] # SSBmsy is the equilibrium SSB at Fmsy
SSBmsy

# 6)
MSY <- DF$yield_f[which(DF$yield_f == max(DF$yield_f))] # MSY is the yield at Fmsy
MSY

```

The MSY reference points are plotted below the the yield curves:

```{r, warning = F, fig.show="hold", out.width="75%"}

# Plot Yield vs. Equilibrium SSB with reference lines at SSBmsy and MSY
ggplot(DF) + geom_point(mapping=aes(y=yield_f,x=eq_ssb_f)) + theme_classic() + 
  labs(x="SSB (kt)",y="Yield (kt)") +
  scale_x_continuous(limits = c(0,10), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,0.5), expand = c(0, 0)) +
  geom_vline(xintercept=SSBmsy, linetype=2, color = "red") +
  geom_hline(yintercept=MSY, linetype=2, color = "red") 
```

<i><b>Figure 3. Equilibrium yield vs. equilibrium SSB over a range of $F$ values. MSY is the maximum yield (horizontal dashed line) and occurs at SSB = $SSB_\text{MSY}$ (vertical dashed line)</i></b> 

```{r, warning = F, fig.show="hold", out.width="75%"}

# Plot Yield vs. F with reference lines at Fmsy and MSY
ggplot(DF) + geom_point(mapping=aes(y=yield_f,x=f)) + theme_classic() + 
  labs(x="F",y="Yield (kt)") +
  scale_x_continuous(limits = c(0,3), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,0.5), expand = c(0, 0)) +
  geom_vline(xintercept=Fmsy, linetype=2, color = "red") +
  geom_hline(yintercept=MSY, linetype=2, color = "red") 
```

<i><b>Figure 4. Equilibrium yield vs. equilibrium $F$ over a range of $F$ values. MSY is the maximum yield (horizontal dashed line) and occurs at F = $F_\text{MSY}$ (vertical dashed line)</i></b> 

### Footnote: Ricker stock-recruit equations

The Ricker stock-recruit function is defined by two parameters $\alpha$ and $\beta$. The $\alpha$ parameter represents the maximum juvenile survival rate and has exactly the same interpretation as in the Beverton-Holt model. However, the Ricker model can allow for "over-compensation" (lower survival at high stock sizes), so the $\beta$ parameter is not equivalent to that in the Beverton-Holt equation.

\begin{equation*}
\tag{7a}
  R = \alpha SSBe^{-\beta SSB}
\end{equation*}

As for Equation 8, we can re-express this function in terms of spawning biomass per recruit:

\begin{equation*}
\tag{8a}
R_0 = \frac{ln(\alpha \phi_{E0})}{\beta \phi_{E0}} 
\end{equation*}

(Walters and Martell 2004, Box 3.1). We assume that the model has estimated the leading parameters $R_{0}$ and steepness $h$. The $\alpha$ and $\beta$ parameters are then derived thus :

\begin{equation*}
\tag{9a}
  \alpha = \frac{5h^{5/4}}{\phi_{E0}}
\end{equation*}

Once we have an estimate of \alpha, we can obtain \beta by rearranging Equation 8a:

\begin{equation*}
\tag{10a}
  \beta = \frac{ln(\alpha \phi_{E0})}{R_0 \phi_{E0}} 
\end{equation*}

## Note about Salmonids

The above calculations apply to iteroparous species, where we need to account for  mortality of mature fish as they age and spawn through the years. We do not have to account for this for semelparous species, such as some salmonids (e.g., Pacific salmon), which spawn only once in their lifetime. For these species, the population can be modeled using only the stock-recruit relationship,  often the Ricker model (Ricker 1954), under the assumption that the estimate of recruits is equivalent to the spawners in the subsequent generation, in the absence of fishing. MSY-based reference points can be estimated directly from the alpha and beta parameters of the Ricker or Beverton-Holt stock-recruit models (see Table 7.2, Hilborn and Walters 1992), Scheuerell (2016) for $S_\text{MSY}$ and $U_\text{MSY}$, and Holt (2009) and Holt et al. (2009) for $S_{gen}$). In Canada, reference points for Pacific salmon populations are frequently based on assessments which rely on various metrics to assign status. See Pestal et al. (2022), previous Wild Salmon Policy assessments (e.g., Grant et al. 2020; DFO 2015a, 2016), and recent papers on Limit Reference Points (LRPs) for Pacific salmon major stocks (Holt et al. In press a; In press b; DFO 2022a). 

For Atlantic salmon, an iteroparous species, reference points are typically based on the egg deposition that results in less than an acceptably low probability that the realized smolt production from freshwater would be less than 50 per cent of the estimated maximum recruitment (DFO 2015b; 2022b). We do not have time to explore estimation of salmonid reference points in these webinars but, in addition to the previous references, see Chaput et al. (2012), Portley and Geiger (2014), Grant et al. (2011; 2013), and Holt and Michielsens (2020) for more information. 

## References

Forrest, R.E., Martell, S.J.D., Melnychuk, M.C., and Walters, C.J. 2008. An age-structured model with leading management parameters, incorporating age-specific selectivity and maturity. Canadian Journal of Fisheries and Aquatic Sciences, 65(2): 286-296. https://doi.org/10.1139/cjfas-2012-0372

Forrest, R.E., McAllister, M.K., Dorn, M., Martell, S.J.D, and Stanley, R. 2010. Hierarchical Bayesian estimation of productivity and reference points for Pacific rockfishes (Sebastes spp.) under alternative assumptions about the stock-recruit function. Canadian Journal of Fisheries and Aquatic Sciences, 67(10): 1611-1634. https://doi.org/10.1139/F10-077

Hilborn R. and Walters, C.J. 1992. Quantitative fisheries stock assessment: choice, dynamics and uncertainty. Chapman and Hall, New York. 570 pp.

Martell, S.J.D., Pine. W.E. III, and Walters, C.J. 2008. Parameterizing age-structured models from a fisheries management perspective. Canadian Journal of Fisheries and Aquatic Sciences, 65:1586-1600. https://doi.org/10.1139/F08-055

Ricker, W.E. 1954. Stock and Recruitment. Journal of the Fisheries Research Board of Canada 11(5), 559–623.

Schnute, J.T. and Kronlund, A.R. 1996. A management oriented approach to stock recruitment analysis. Canadian Journal of Fisheries and Aquatic Sciences 53, 1281-1293. https://doi.org/10.1139/f96-069

Walters, C.J. and Martell, S.J.D. 2004. Fisheries Ecology and Management. Princeton University Press, Princeton, 399 pp.

**Salmon references**

Chaput, G., Cass, A., Grant, S., Huang, A.-M., & Veinott, G. 2012. Considerations for Defining Reference Points for Semelparous Species, with Emphasis on Anadromous Salmonid Species Including Iteroparous Salmonids. DFO Canadian Science Advisory Secretariat Research Document, 2012/146, v + 48 p.

DFO. 2015a. Wild Salmon Policy Biological Status Assessment for Conservation Units of Interior Fraser River Coho Salmon (*Oncorhynchus kisutch*). DFO Canadian Science Advisory Secretariat Science Advisory Report, 2015/022.

DFO. 2015b. Development of Reference Points for Atlantic Salmon (Salmo Salar) That Conform to the Precautionary Approach. DFO Can. Sci. Advis. Sec. Sci. Advis. Rep., 2015/058.

DFO. 2016. Integrated Biological Status of Southern British Columbia Chinook Salmon (*Oncorhynchus tshawytscha*) Under the Wild Salmon Policy. DFO Canadian Science Advisory Secretariat Science Advisory Report, 2016/042.

DFO. 2022a. Methodologies and Guidelines for Defining Limit Reference Points for Pacific Salmon. DFO Can. Sci. Advis. Sec. Sci. Advis. Rep., 2022/030.

DFO. 2022b. Definition of Precautionary Approach Reference Points for Atlantic Salmon, DFO Gulf Region. DFO Can. Sci. Advis. Sec. Sci. Advis. Rep, 2022/027.

Grant, S.C.H., MacDonald, B. L., Cone, T. E., Holt, C. A., Cass, A., Porszt, E. J., Hume, J. M. B., & Pon, L. B. 2011. Evaluation of Uncertainty in Fraser Sockeye (*Oncorhynchus Nerka*) Wild Salmon Policy Status Using Abundance and Trends in Abundance Metrics (DFO Can. Sci. Advis. Sec. Res. Doc. No. 2011/087; p. viii + 183).

Grant, S.C.H., & Pestal, G. 2013. Integrated Biological Status Assessment under the Wild Salmon Policy Using Standardized Metrics and Expert Judgement: Fraser River Sockeye Salmon (*Oncorhynchus nerka*) Case Studies. DFO Can. Sci. Advis. Sec. Res. Doc., 2012/106, 137.

Grant, S.C.H., Holt, C.A., Pestal, G., Davis, B.M., & MacDonald, B.L. 2020. The 2017 Fraser Sockeye Salmon (*Oncorhynchus nerka*) Integrated Biological Status Re-Assessments Under the Wild Salmon Policy Using Standardized Metrics and Expert Judgment. DFO Can. Sci. Advis. Sec. Res. Doc., 2018/017, vii + 211 p.

Hilborn, R., Walters, C. J. 1992. Quantitative Fisheries Stock Assessment: Choice, Dynamics and Uncertainty. Springer Science & Business Media.

Holt, C.A. 2009. Evaluation of Benchmarks for Conservation Units in Canada’s Wild Salmon Policy: Technical Documentation (DFO Canadian Science Advisory Secretariat Science Reserach Document No. 2009/059; p. x + 50).

Holt, C.A., Michielsens, C.G.J. 2020. Impact of Time-Varying Productivity on Estimated Stock-Recruitment Parameters and Biological Reference Points. Canadian Journal of Fisheries and Aquatic Sciences, 77(5), 836–847.

Holt, C.A., Cass, A., Holtby, B., & Riddell, B. 2009. Indicators of Status and Benchmarks for Conservation Units in Canada’s Wild Salmon Policy. Canadian Science Advisory Secretariat Science Research Document, 2009/058, 82.

Holt, C.A., Holt, K., Warkentin, L., Wor, C., Connors, B., Grant, S., Huang, A.-M., Marentette, J. (In press a). Guidelines for Defining Limit Reference Points for Pacific Salmon Stock Management Units. Canadian Science Advisory Secretariat Research Document, 202n/nnn.

Holt, K.R., Holt, C.A., Warkentin, L., Wor, C., Davis, B., Arbeider, M., Bokvist, J., Crowley, S., Grant, S., Luedke, W., McHugh, D., Picco, C.,  Van Will, P. (In press b). Case Study Applications of LRP Estimation Methods to Pacific Salmon Stock Management Units. Canadian Science Advisory Secretariat Research Document, 202n/nnnn.

Pestal, G., MacDonald, B.L., Grant, S. C. H., & Holt, C. A. 2022. State of the Salmon: Rapid Status Assessment Approach for Pacific Salmon under Canada’s Wild Salmon Policy. Can. Tech. Rep. Fish. Aquat. Sci, In review.

Portley, N., & Geiger, H.J. 2014. Limit Reference Points for Pacific Salmon Fisheries. North American Journal of Fisheries Management, 34(2), 401–410. https://doi.org/10.1080/02755947.2014.882453

Scheuerell, M.D. 2016. An Explicit Solution for Calculating Optimum Spawning Stock Size from Ricker’s Stock Recruitment Model. PeerJ, 4, e1623. https://doi.org/10.7717/peerj.1623

