---
title: "TESA Reference Points 101"
output: html_document
author: "Tim Barrett and Robyn Forrest"
date: "November 15 and 22, 2022"
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = T, warnings= F)
```

## Exercise 1: MSY reference points using a Schaefer Surplus Production Model

***DISCLAIMER: These are illustrative, deterministic models and should not be used for stock assessment***

**Run the R chunks in Rstudio using the green arrow in each chunk. Run individual lines in the chunks using ctrl+enter. If you want to explore further, you can add your own R code inside any of the chunks, or add your own chunks using ctrl+alt+i. Click on the Knit button in Rstudio to create the document.**

In this exercise we'll explore the idea of projecting the population forward under equilibrium conditions using a surplus production model. By the end of the exercise you will learn:

* Fitting a basic surplus production model and calculating reference points based on Maximum Sustainable Yield from the Schaefer model.

* Projecting under equilibrium conditions.

* What is meant by the term "equilibrium reference points".

## 1. Surplus production model with swordfish data

A time series of catch (1950-2005) and survey index (1963-1970 and 1975-2005) are provided for a swordfish stock in dataframe **DF** (source: [openMSE](https://openmse.com/)). 

```{r, echo=F,warning=F,figures-side, fig.show="hold", out.width="50%"}
DF <- read.csv("swordfish.csv")

ggplot(DF) + geom_path(mapping=aes(y=Catch,x=Year), size=1.5) + theme_classic() +
  scale_x_continuous(limits = c(1950,2005), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,22000), expand = c(0, 0))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=12))
ggplot(DF) + geom_path(mapping=aes(y=Index,x=Year), colour="blue", size=1.5) + theme_classic() +
  scale_x_continuous(limits = c(1950,2005), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,1200), expand = c(0, 0))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold"))
```

<i><b>Figure 1. Time series plots of historical catch and the index (1950-2005)</i></b> 

```{r}
tail(DF) # Preview data
```


We would like to estimate model parameters and MSY-based reference points for this stock. To do this, we first need to fit a population model to the data to estimate the biomass, fishing mortality rate and model parameters. We'll start with a simple Schaefer, or Surplus Production (SP), model (Schaefer 1954). The Schaefer model is: 

\begin{equation*}
\tag{1}
B_{t+1} = B_{t} + r B_{t} \left(1-\frac{B_{t}}{K}\right) - C_{t}
\end{equation*}

where $B$ is biomass, $r$ is the intrinsic rate of population growth, $K$ is the carrying capacity, $C$ is the catch, and $t$ is a discrete time period (one year in this case). $B_\text{MSY}$ is then calculated as $K/2$ and $F_\text{MSY}$ is calculated as $r/2$, where $F$ is the instantaneous fishing mortality rate (Hilborn and Walters 1992, Ch. 8; see slides **XXX** of the PPT).

We could fit the model using various software packages, or write the code ourselves in R or other languages like ADMB, stan or TMB. Here we use SAMtool, an R package that is part of the openMSE software. When you install the openMSE R package, you will automatically get SAMtool. 

```{r,warning=F,message=F}
library(SAMtool) # Used to fit the surplus production model using function SP()

# Create a data object D with the Catch, Index, Years, and coefficient of variation for the Index.
D <- new('Data')
D@Name <- "swordfish"
D@Cat <- matrix(DF$Catch,nrow=1)
D@Year <- 1950:2005
D@Ind <- matrix(DF$Index,nrow=1)
D@CV_Ind <- matrix(0.3, nrow=1, ncol=length(D@Year)) # This is a coefficient of variation on the index. Assumed 0.3 but doesn't influence reference points.

# Run the model
Model <- SP(Data = D)

# Extract the parameter estimates from the model
K <- Model@TMB_report$K
r <- Model@TMB_report$r
BMSY <- K/2
FMSY <- r/2

# Add the biomass and BMSY estimates to the DF object
DF$Model_biomass <- Model@B[1:length(D@Year)]
DF$BMSY <- BMSY

# Plot
ggplot(DF) + 
  geom_path(mapping=aes(y=Model_biomass,x=Year, colour="Model biomass"), size=1.5)+
  geom_path(mapping=aes(y=BMSY,x=Year, colour="BMSY"), lty=2, size=1.5)+
  scale_color_manual(name="",
                     breaks=c("Model biomass","BMSY"),
                     values=c("Model biomass"="purple", "BMSY"="black"))+
  theme_classic() +
  xlim(1950,2005) +
  ylim(0,110000)+
  ylab("Model Biomass")+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold"))

```

<i><b>Figure 2. Time series plot of model estimated biomass over the historical time period with a horizontal line at $B_\text{MSY}$</i></b> 


## 2. Numerical projection under constant fishing mortality rates

To help understand the concept of equilibrium reference points, let's look at projecting the population forward for 100 years under a constant fishing mortality rate $F$. We'll start with the most recent fishing mortality rate (terminalF) to show that the population eventually reaches an equilibrium, or steady, state. In the real world we would interpret this as a long-term average rather than exactly the same value every year.

```{r,warning=F}

# Define indexes
nyr <- length(DF$Year) # number of historical years
pyr <- 100 # number of projected years
fpyr <- nyr+pyr # index of final projection year
pyrs <- (DF$Year[nyr]+1):(DF$Year[nyr]+pyr)
ayrs <- (DF$Year[1]):(DF$Year[nyr]+pyr)

# Make a dataframe of historical data and biomass
P <- data.frame("Year"=DF$Year, 
    "Biomass"=DF$Model_biomass, 
    "Catch"=DF$Catch,
    "U"=DF$Catch/DF$Model_biomass,
    "FM"=-log(1-(DF$Catch/DF$Model_biomass))) 
     
# Add cells for the projection years
P[fpyr,] <- NA

# Final year biomass, catch and U
terminalB <- P$Biomass[nyr]
terminalC <- P$Catch[nyr]
terminalU <- terminalC/terminalB
terminalF <- -log(1-terminalU)

# Populate the projection years
P$Year[(nyr+1):fpyr] <- pyrs
P$FM[(nyr+1):fpyr] <- terminalF

for(y in (nyr+1):fpyr)
  { # loop over projection years
    lastB <- P[(y-1),"Biomass"]
    lastC <- P[(y-1),"Catch"]
    
    P[y,"Biomass"] <- lastB + r*lastB*(1-lastB/K)-lastC # Projected Biomass from the Schaefer model
    P[y,"Catch"] <- lastB*terminalF # Projected Catch from the Schaefer model
}

# Plot     
ggplot(P) + 
  geom_path(mapping=aes(y=Biomass,x=Year, colour="Biomass"), size=1.5)+
  geom_path(mapping=aes(y=Catch,x=Year, colour="Catch"), size=1.5)+
  scale_color_manual(name="",
                     breaks=c("Biomass","Catch"),
                     values=c("Biomass"="purple", "Catch"="black"))+
  geom_vline(xintercept= P$Year[nyr], lty=2)+
  theme_classic() +
  ylab("Biomass and Catch")+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold")) 

```

<i><b>Figure 3. Time series plot of biomass and catch for the historical time period and projected over 100 years using the fishing mortality rate from the terminal year. The vertical dashed line shows the beginning of the projection period under constant $F$. </i></b> 

**Figure 3** shows the historical population biomass to the left of the dashed vertical line, and the projected biomass to the right. The population reached equilibrium under the constant fishing mortality rate quite quickly. Just to extend this a bit, let's do the same with four different fishing mortality rates: $F = 0$, $F =$ terminalF, $F = F_\text{MSY}$, and a high fishing mortality rate $F = F_{crash}$ that would crash the population. Note that $F_\text{MSY} = r/2$ from the fitted model.

```{r,warning=F}

library(reshape2) # Needed for the melt function

# make a vector of values for U and an array P2 for results
f <- c(0,terminalF, FMSY, r*0.96)
P2 <- array(NA,dim=c(100,length(f),2),dimnames = list(1:100,f,c("Biomass","Catch")))

for(i in 1:length(f)){ #loop over f
  for(y in 1:pyr){     #loop over years
    if(y==1){          #get terminal year (2005) catch and biomass to begin projections for year 1
      lastB <- terminalB
      lastC <- terminalC
    } else {
      lastB <- P2[(y-1),i,"Biomass"]
      lastC <- P2[(y-1),i,"Catch"]
    }
    P2[y,i,"Biomass"] <- lastB + r*lastB*(1-lastB/K)-lastC # Estimated Biomass from the Schaefer model
    P2[y,i,"Catch"] <- lastB*f[i] # Estimated Catch from the Schaefer model
  }
}

# Plot the biomass
pB <- P2[,,1] %>% 
  as.data.frame() %>% 
  rename("0" = 1,
         Fterminal = 2,
         FMSY = 3,
         Fcrash = 4) %>% 
  mutate("Year" = pyrs) %>% 
  melt(id="Year",variable="FM", value.name="Biomass")

# Plot biomass under each F
ggplot(pB, aes(Year, Biomass)) + 
  geom_line(aes(color = FM), size=1.5) +
  geom_line(data=DF,aes(y=Model_biomass,x=Year), colour="purple", size=1.5)+
  geom_vline(xintercept=ayrs[nyr], lty=2)+
  theme_classic() +
  ylab("Biomass")+xlab("Year")+
  ylim(0,110000)+
  xlim(ayrs[1],ayrs[length(ayrs)])+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold")) 

```

<i><b>Figure 4. Time series plot of biomass for the historical time period and 100 projected years for four different constant fishing mortality rates.  The vertical dashed line shows the beginning of the projection period under constant $F$. </i></b> 

Fishing constantly at any fishing mortality rate results in the stock coming to some equilibrium biomass $B$, ranging between $B = 0$ and $B = K$. The Schaefer model predicts that the stock will eventually come to carrying capacity $K$ under no fishing, and to $B_\text{MSY}$ under a constant fishing mortality rate of $F_\text{MSY}$, as shown in the plot.

We can see this by comparing final year equilibrium values from the above numerical exercise with the analytical predictions of the Schaefer model (i.e., $B_\text{MSY} = K/2$ and $F_\text{MSY} = r/2$). The numerical and analytical solutions for $K$ and $B_\text{MSY}$ are identical.

```{r,warning=F}

# Just get the Biomass from the P2 array
# Column headers are fishing mortality rate f
pB <- P2[,,"Biomass"] %>% 
  as.data.frame()
tail(pB)

# Final Year Biomass under no fishing
Beq_0 <- pB[pyr,1] # Column 1 is f = 0
# Compare to K
Beq_0 # Numeric solution
K     # Analytical solution

# Final Year Biomass under Umsy
Beq_fmsy <- pB[pyr,3] # Column 3 is f = Fmsy
Beq_fmsy # Numerical solution
K/2    # Analytical solution

```

This is what is meant by "equilibrium reference points", i.e., if we run out a model for a long time under constant fishing mortality rate and constant conditions, it will "equilibrate" at the analytically-predicted values. We will see in the next exercise that there is no analytical solution for $F_\text{MSY}$ in age-structured models, but the idea that equilibrium reference points represent the results of running out a model for a long time under constant fishing mortality rates is the same (although there are fortunately short-cut numerical methods so we don't have to do this).

Of course, in the real world there are never constant conditions. Equilibrium reference points such as $B_\text{MSY}$ represent some long-term average with annual fluctuations around this average. This assumes there are no directional trends or major shifts in model parameters or environmental conditions. The problem of non-stationarity becomes much more complex with statistical catch-at-age models, which may have trends or shifts in parameters such as growth, natural mortality or selectivity-at-age, or in observed population data such as weight-at-age. Calculating reference points under these conditions requires some extra assumptions (such as whether to average time-varying variables and over what period), which are often quite subjective. These choices can be impactful in terms of interpretation of reference points and stock status.  Impacts of time-varying variables on reference points and advice is an active area of research (e.g., Haltuch et al. 2009; Holt and Michielsens 2020; Zhang et al. 2020). In this webinar we are trying to illustrate the basic concepts. We will touch on non-stationarity in Exercise 5, but will not have time for detailed analysis.

## 3. Understanding 'Maximum Sustainable Yield'

We've seen that $B_\text{MSY}$ is an analytical property of the Schaefer model $\left(K/2\right)$, which can be reproduced numerically by running out the model under a constant fishing mortality rate of $F_\text{MSY}$. But where did $F_\text{MSY}$ come from? Next, we'll repeat Step 2 over a range of fishing mortality rates from 0 to 1 in increments of 0.001 and show that the constant fishing mortality rate $F$ that maximizes long-term yield is the same as the estimate of $F_\text{MSY}$ from Section 1.

The two-parameter Schaefer model produces a symmetrical yield curve. More complex, non-symmetrical versions of this model exist (e.g., the Pella-Tomlinson model; Hilborn and Walters 1992) and, indeed, similar yield curves are produced by age-structured models. The important concept from this exercise is that $F_\text{MSY}$ is the long-term fishing mortality rate that would maximize yield in the long-term under stable conditions.

In the exercise below, we identify $F_\text{MSY}$ numerically by running out the model under a sequence of fishing mortality rates and identifying the value of $F$ that maximizes yield in the long term.

```{r,warning=F}
# Test values
# r <- 0.25
# K <- 100
# FMSY <- r/2

# make a sequence of values for U and an array P3 for results
f <- seq(0,1,0.0005)
yrs <- 1000
P3 <- array(0,dim=c(yrs,length(f),2),dimnames = list(1:yrs,f,c("Biomass","Catch")))

for(i in 1:length(f)){
  for(y in 1:yrs){
    if(y==1){
      lastB <- terminalB
      lastC <- terminalC
     } else {
      lastB <- P3[(y-1),i,"Biomass"]
      lastC <- P3[(y-1),i,"Catch"]
    }

    P3[y,i,"Biomass"] <- lastB + r*lastB*(1-lastB/K)-lastC #Biomass
    if(P3[y,i,"Biomass"]<0.000001) P3[y,i,"Biomass"] <- 0.000
    # Get the catch for next year
    P3[y,i,"Catch"] <- P3[y,i,"Biomass"] * f[i]  # Catch = qEB, but qE = F (assume q=1)
  }
}

# use last year for equilibrium 
EX3 <- data.frame(FM=f,
                  Yield = P3[yrs,,2],
                  equil_B = P3[yrs,,1])

#test
maxY  <- max(EX3$Yield)
fmaxY <- EX3$FM[which(EX3$Yield==maxY)]
fmaxY
round(FMSY,4)

```

As you can see, the analytical solution for $F_\text{MSY}$, given by $r/2$ is identical to the numerical solution, which is the fishing mortality rate that produced the maximum final year yield (fmaxY).


```{r}
# Plot

ggplot(EX3)+
  geom_path(mapping=aes(y=Yield ,x=FM), size=1.5) + 
  theme_classic() +
  scale_x_continuous(limits = c(0,1), expand = c(0,0)) +
  scale_y_continuous(limits = c(0,20000), expand = c(0,0)) +
  # add vertical line at Fmsy estimated from model parameter r
  geom_vline(xintercept = r/2, linetype = 2, colour = "blue", size=1.25)+
  xlab("Fishing Mortality")+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.text.y = element_text(size=12))+
  theme(axis.title.x = element_text(size=12, face="bold"))+
  theme(axis.title.y = element_text(size=12, face="bold"))

```

<i><b>Figure 5. Plot of yield vs. Fishing Mortality rate ($F$) with a vertical line at $F_\text{MSY}$ where yield is maximized</i></b> 

**Figure 5** shows the final year catch (i.e., yield) under each of the tested fishing mortality rates. It confirms that long-term yield is maximized by fishing at $F_\text{MSY}$, indicated by the blue dashed line.

Note: The figure above (an inverted parabola) suggests that we could use calculus to find the fishing mortality rate that maximizes yield (i.e., the derivative of yield with respect to $F$ is zero; See slides **XXX** of the powerpoint). This is indeed how most numerical searches for $F_\text{MSY}$ work. For the purposes of these exercises, however, we used a simpler grid search. This most clearly illustrates the concept of maximizing yield, but please be aware that this is the least efficient means of finding $F_\text{MSY}$!

## References

Haltuch, M.A., Punt, A.E., & Dorn, M.W. 2009. Evaluating the estimation of fishery management reference points in a variable environment. Fisheries Research, 100(1), 42–56. https://doi.org/10.1016/j.fishres.2009.03.001 

Hilborn R. and Walters, C.J. 1992. Quantitative fisheries stock assessment: choice, dynamics and uncertainty. Chapman and Hall, New York. 570 pp.

Holt, C.A., & Michielsens, C.G.J. 2020. Impact of time-varying productivity on estimated stock-recruitment parameters and biological reference points. Canadian Journal of Fisheries and Aquatic Sciences. https://cdnsciencepub.com/doi/10.1139/cjfas-2019-0104

Schaefer, M.B. 1954. Some aspects of the dynamics of the populations important to the management of the commercial marine fisheries. Inter-American Tropical Tuna Commission, 1: 27-56.

Zhang, F., Regular, P.M., Wheeland, L., Rideout, R.M., & Morgan, M.J. (2020). Accounting for non-stationary stock–recruitment relationships in the development of MSY-based reference points. ICES Journal of Marine Science, fsaa176. https://doi.org/10.1093/icesjms/fsaa176

**Further reading**

Cadrin, S.X. 2012. Unintended consequences of MSY proxies for defining overfishing. ICES Evolution of management frameworks to prevent overfishing. ICES CM 2012/L:23. 

Larkin, P. A. 1977. An epitaph for the concept of maximum sustained yield. Transactions of the American Fisheries Society 106(1), 1-11. [https://doi.org/10.1577/1548-8659(1977)106<1:AEFTCO>2.0.CO;2](https://doi.org/10.1577/1548-8659(1977)106<1:AEFTCO>2.0.CO;2)

Mace, P.M., 2001. A new role for MSY in single-species and ecosystem approaches to fisheries
stock assessment and management. Fish and Fisheries 2(1), 2-32.  https://doi.org/10.1046/j.1467-2979.2001.00033.x

Punt, A.E. and Smith, A.D.M. 2001. The gospel of Maximum Sustainable Yield in fisheries management: birth, crucifixion and reincarnation. In: Conservation of Exploited Species (eds. J.D. Reynolds, G.M. Mace, K.H. Redford & J.G. Robinson), pp. 41-66. Cambridge University Press, Cambridge.

Sainsbury, K. 2008. Best Practice Reference Points for Australian Fisheries. Australian Fisheries Management Authority Report R2001/0999.
