---
title: "TESA Reference Points 101"
output: html_document
author: "Tim Barrett, Irene Andrushchenko, and Robyn Forrest"
date: "November 15 and 22, 2022"
---

```{r setup, include=FALSE}
library(ggplot2)
library(kableExtra)
library(viridis)
library(reshape2)
library(cowplot)
source("source_ftns.r")
knitr::opts_chunk$set(echo = T, warnings= F)
```

## Exercise 5B:
## Reference points using an age-structured model with time-varying $M$

***DISCLAIMER: These are illustrative, deterministic models and should not be used for stock assessment***

**Run the R chunks in Rstudio using the green arrow in each chunk. Run individual lines in the chunks using ctrl+enter. If you want to explore further, you can add your own R code inside any of the chunks, or add your own chunks using ctrl+alt+i. Click on the Knit button in Rstudio to create the document.**

This exercise is similar to Exercise 5A but with time-varying $M$.

Here we look at an operating model (OM) where $M$ for the post-recruit ages is time-varying. The objective of this exercise is to explore the influence of time-varying $M$ on $B_\text{MSY}$ and $B_0$. 

The OM was conditioned on age-composition data for the catch and a survey index for a pollock stock (time series of 39 years). An R object is provided for the OM that contains the relevant data for the exercise. The model was conditioned using a Beverton-Holt stock-recruit relationship (SRR) with an initial assumed $h$ of 0.65 and time-varying $M$. Some simplifying assumptions are made here regarding $M$ over time, i.e., juvenile (age 0-3) $M$-at-age is a constant at 0.2, and adult (age 4+) $M$-at-age was estimated in specific time blocks which we will see below. We assume  that future selectivity-at-age is equal to the maturity-at-age.

The model-estimated Beverton-Holt SRR parameters $\alpha$ and $\beta$ are provided.

We begin by reading in the data and plotting $M$-at-age over the historical time period.

```{r,warning=F,message=F}
datalist <- readRDS("Ex5data.rda")
# The DF dataframe has the data needed for this exercise
# DF contains columns with Year, model estimated SSB in t, model estimated age-0 recruitment (thousands), the Beverton-Holt a and b parameters, and the age (0 to 13+), weight-at-age (waa in kg), proportion mature-at-age (mat), and assumed future selectivity-at-age (sel = mat).
DF <- datalist$DF
head(DF)
# M is time-varying and a matrix of dimension 14 x 39 is provided with the weights-at-age 0 to 13+ for 39 years
MAA <- datalist$MAA
MAA[,c(1,15,25,39)] # Preview M-at-age with one year in each time block
```

Plot of $M$-at-age over time:

```{r,warning=F}
MAADF <- melt(MAA)
colnames(MAADF)<-c('Age','Year','M')

ggplot(MAADF, aes(y=Age, x=Year)) + geom_tile(aes(fill=M)) + scale_fill_viridis() + theme_bw() +
  scale_y_continuous(breaks = 1:13, limits = c(0,14), expand = c(0,0)) + 
  scale_x_continuous(breaks = seq(0,40,10), limits = c(0,40), expand = c(0,0))
```

<i><b>Figure 1. Model-assumed $M$-at-age time series over 39 historical years. Age 13 is a plus group</i></b> 
<br>

## Influence of changes in age 4+ $M$-at-age on $SSB_0$, and $SSB_\text{MSY}$

We explore the influence of using $M$-at-age from one year in each time block:

a\right) year 1 of the time series, 

b\right) year 15 of the time series, 

c\right) year 25 of the time series, or

d\right) year 39 of the time series

on estimates of the unfished SSB-per-recruit ($\phi_{E0}$), $SSB_0$, and $SSB_\text{MSY}$.

```{r,warning=F,message=F}
# unfished SSB-per-recruit (kg/rec)
phi0_a <- sum(survivorship_F(f=0,M=MAA[,1],n_ages=14)*DF$waa[1:14]*DF$mat[1:14])
phi0_b <- sum(survivorship_F(f=0,M=MAA[,15],n_ages=14)*DF$waa[1:14]*DF$mat[1:14])
phi0_c <- sum(survivorship_F(f=0,M=MAA[,25],n_ages=14)*DF$waa[1:14]*DF$mat[1:14])
phi0_d <- sum(survivorship_F(f=0,M=MAA[,39],n_ages=14)*DF$waa[1:14]*DF$mat[1:14])

c(phi0_a, phi0_b, phi0_c, phi0_d)

# SSB0
SSB0_a <- (DF$BHa[1]*phi0_a-1)/DF$BHb[1]
SSB0_b <- (DF$BHa[1]*phi0_b-1)/DF$BHb[1]
SSB0_c <- (DF$BHa[1]*phi0_c-1)/DF$BHb[1]
SSB0_d <- (DF$BHa[1]*phi0_d-1)/DF$BHb[1]

c(SSB0_a, SSB0_b, SSB0_c, SSB0_d)

```

**Figure 2** shows the SRR with the lines through the origin with slope $\frac{1}{\phi_{E0}}$ for each $M$ that intersect the SRR at each value of $SSB_{0}$.


```{r,warning=F}
ggplot(DF)+
  geom_point(mapping=aes(y=Rec,x=SSB)) + 
  theme_classic() +
  labs(x="SSB (t)",y="Age 0 Recruitment in thousands") +
  scale_x_continuous(limits = c(0,120000), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,40000), expand = c(0, 0)) +
  geom_function(fun=function(x) (DF$BHa[1]*x/(1+DF$BHb[1]*x))) +
  geom_function(fun=function(x) (1/phi0_a*x),colour="midnightblue",linetype=2) +
  geom_function(fun=function(x) (1/phi0_b*x),colour="yellow2") + geom_function(fun=function(x) (1/phi0_b*x),linetype=2) +
  geom_function(fun=function(x) (1/phi0_c*x),colour="green2",linetype=2) +
  geom_function(fun=function(x) (1/phi0_d*x),colour="dodgerblue2",linetype=2) +
  geom_vline(xintercept=SSB0_a,color="midnightblue") +
  geom_vline(xintercept=SSB0_b,color="yellow2") + geom_vline(xintercept=SSB0_b,linetype=3) +
  geom_vline(xintercept=SSB0_c,color="green2") +
  geom_vline(xintercept=SSB0_d,color="dodgerblue2") 

```

<i><b>Figure 2. Stock-recruit pairs and model-estimated stock-recruit relationship (SRR) with lines through the origin with slopes $\frac{1}{\phi_{E0}}$ with colours matching $M$ in Figure 1 that intersect the SRR at different $B_0$ values (vertical lines)</i></b> 
<br>

Next we look at the influence of the change in $M$ on $B_\text{MSY}$ and $F_\text{MSY}$, where the `MSYcalc` function from "source_ftns.r" is used to calculate $B_\text{MSY}$.

```{r,warning=F}

RP_a <- MSYcalc(M=MAA[,1],waa=DF$waa[1:14],mat=DF$mat[1:14],sel=DF$sel[1:14],a=DF$BHa[1],b=DF$BHb[1])
RP_b <- MSYcalc(M=MAA[,15],waa=DF$waa[1:14],mat=DF$mat[1:14],sel=DF$sel[1:14],a=DF$BHa[1],b=DF$BHb[1])
RP_c <- MSYcalc(M=MAA[,25],waa=DF$waa[1:14],mat=DF$mat[1:14],sel=DF$sel[1:14],a=DF$BHa[1],b=DF$BHb[1])
RP_d <- MSYcalc(M=MAA[,39],waa=DF$waa[1:14],mat=DF$mat[1:14],sel=DF$sel[1:14],a=DF$BHa[1],b=DF$BHb[1])

RES <- data.frame(M=c(MAA[14,c(1,39,25,15)]),phi_E0=round(c(phi0_a,phi0_d,phi0_c,phi0_b),2),
                  SSB0=round(c(SSB0_a, SSB0_d, SSB0_c, SSB0_b),0),
                  SSBmsy=round(c(RP_a$SSBmsy, RP_d$SSBmsy, RP_c$SSBmsy, RP_b$SSBmsy),0),
                  Fmsy=c(RP_a$Fmsy, RP_d$Fmsy, RP_c$Fmsy, RP_b$Fmsy))
```

```{r,warning=F,message=F}
# Plot PhiE0, SSB0, SSBMSY and FMSY against age 4+ M

g1 <- ggplot(RES)+
  geom_point(mapping=aes(y=phi_E0,x=M), size =3, colour="dodgerblue2") +
  geom_path(mapping=aes(y=phi_E0,x=M), colour="dodgerblue2")+
  theme_classic()+
  labs(x="Age 4+ M",y="phiE0") +
  ylim(0,1.1*phi0_a)+
  theme(axis.text.x = element_text(size=12)) +
  theme(axis.text.y = element_text(size=12)) +
  theme(axis.title.x = element_text(size=12, face="bold")) +
  theme(axis.title.y = element_text(size=12, face="bold"))
g2 <- ggplot(RES)+
  geom_point(mapping=aes(y=SSB0,x=M), size =3, colour="dodgerblue2") + 
  geom_path(mapping=aes(y=SSB0,x=M), colour="dodgerblue2")+
  theme_classic()+
  labs(x="Age 4+ M",y="SSB0")+
  ylim(0,1.1*SSB0_a)+
  theme(axis.text.x = element_text(size=12)) +
  theme(axis.text.y = element_text(size=12)) +
  theme(axis.title.x = element_text(size=12, face="bold")) +
  theme(axis.title.y = element_text(size=12, face="bold"))
g3 <- ggplot(RES)+
  geom_point(mapping=aes(y=SSBmsy,x=M), size =3, colour="dodgerblue2") + 
  geom_path(mapping=aes(y=SSBmsy,x=M), colour="dodgerblue2")+
  theme_classic()+
  labs(x="Age 4+ M",y="SSBMSY") +
  ylim(0,1.1*RP_a$SSBmsy)+
  theme(axis.text.x = element_text(size=12)) +
  theme(axis.text.y = element_text(size=12)) +
  theme(axis.title.x = element_text(size=12, face="bold")) +
  theme(axis.title.y = element_text(size=12, face="bold"))
g4 <- ggplot(RES)+
  geom_point(mapping=aes(y=Fmsy,x=M), size =3, colour="dodgerblue2") + 
  geom_path(mapping=aes(y=Fmsy,x=M), colour="dodgerblue2")+
  theme_classic()+
  labs(x="Age 4+ M",y="FMSY")+
  ylim(0,1.1*RP_b$Fmsy)+
  theme(axis.text.x = element_text(size=12)) +
  theme(axis.text.y = element_text(size=12)) +
  theme(axis.title.x = element_text(size=12, face="bold")) +
  theme(axis.title.y = element_text(size=12, face="bold"))

cowplot::plot_grid(g1,g2, g3,g4, ncol=2)

```

<i><b>Figure 3. Relationship between Age 4+ $M$ and $\phi_{E0}$, $SSB_{0}$, $SSB_\text{MSY}$, and $F_\text{MSY}$.</i></b> 
<br>
<br>
<br>
<i><b>Table 1. $\phi_{E0}$, $SSB_0$, $SSB_\text{MSY}$, and $F_\text{MSY}$ for various $M$ (constant $M$ for adult fish)</i></b> 

```{r,echo=F}
RES %>% kbl(align=c(rep('c',4))) %>% kable_styling()
```

We can see the influence of temporal changes in $M$ on the equilibrium reference point calculations in **Table 1** and **Figure 3**. As $M$ increases, $\phi_{E0}$ decreases from reduced survivorship which decreases $SSB_0$. The increased $M$ increases productivity so the stock, in theory, is more resilient to fishing as observed from the higher $F_\text{MSY}$ and lower $SSB_\text{MSY}$. 

In general, a **higher** $M$ results in a **lower** estimate of $B_0$ and $B_\text{MSY}$, and a **higher** $F_\text{MSY}$ (He et al. 2011; Szuwalski and Punt 2012; Punt et al. 2021). However, we advise caution in changing reference points when productivity is changing, as this also implies changing objectives for the stock. See Day 2 presentation for discussion of the management implications of reducing reference points in the presence of increasing $M$. See also Haltuch and Punt (2011) and Punt et al. (2021).

Note that if selectivity is far to the right of maturity, implying that fish have several opportunities to spawn before becoming vulnerable to fisheries, then $F_\text{MSY}$ will be largely insensitive to $M$. This is because, in that situation, the exploitation rate on the <u>vulnerable</u> biomass (which has already spawned) will approach 1 (see Myers and Mertz 1998 for more on this issue). 

# References:

Haltuch, M. A., and Punt, A. E. 2011. The promises and pitfalls of including decadal-scale climate forcing of recruitment in groundfish stock assessment. Canadian Journal of Fisheries and Aquatic Sciences, 68(5), 912–926. https://doi.org/10.1139/f2011-030

He, X., Ralston, S., and MacCall, A.D. 2011. Interactions of age-dependent mortality and selectivity functions in age-based stock assessment models. Fishery Bulletin, 109: 198-216.

Myers, R. A., & Mertz, G. (1998). THE LIMITS OF EXPLOITATION: A PRECAUTIONARY APPROACH. Ecological Applications, 8(sp1), S165–S169. https://doi.org/10.1890/1051-0761(1998)8[S165:TLOEAP]2.0.CO;2

Punt, A.E., Castillo-Jordan, C., Hamel, O.S., Cope, J.M., Maunder, M.N., and Ianelli, J.N. 2021. Consequences of error in natural mortality and its estimation in stock assessment models. Fisheries Research, 233: 105759.

Szuwalski, C., and Punt, A.E. 2012. Identifying research priorities for management under uncertainty: the estimation ability of the stock assessment method used for eastern Bering Sea snow crab (Chionoecetes opilio). Fisheries Research, 134-136, 82-94.